Toward dynamic models of combat or lack thereof in animal mating

### 2016-01-30 ms.tex
ms.tex clean up and exponential event jumping (no more urning)

We consider a population of males competing for mates. The population is one with non-overlapping generations, and within each generation there is a predefined annual cycle. In spring, up to $K$individuals hatch from eggs, grow throughout spring and summer to increase size, and then within a short period, the entire population mates, females lay eggs and everyone dies. The short duration of mating period is such that each male and female only mates once. The population then re-establishes from eggs the following spring.

@@

\subsection{Immature phase}

Within the immature phase, we assume males grow with rate
\begin{equation} \label{eq:growth}
\frac{{\rm d} m}{{\rm d} t} = a \, m^ b
\end{equation}
where $a$ is mass-based rate of growth. Integrating eq. \ref{eq:growth}, the size of individuals at time $t$, having started growing at time $0$ is given by
\begin{equation} \label{eq:growth}
m(t) = \left(m(0)^{1-b} + t a(1-b)\right)^{\frac1{1-b}},
\end{equation}
while the time $\tau$ taken for a male to grow from $m(0) \rightarrow m$ is then
\begin{equation} \label{eq:tau}
\tau(m_0, m) = \frac1{a(1-b)}\left(m ^{1-b} - m_0 ^{1-b}\right).
\end{equation}

@@

\subsection{Searching Males}
When males mature they begin searching for a females nest

The area that a male $m$ with speed trait $v$ and radius trait $r$ can search in one time step is
\begin{equation}
    2r \cdot v
\end{equation}
See Fig. \ref{fig:area searched}.

@@

The mean encounter rate $\varepsilon$ is:
\begin{equation} \label{eq:encounter rate}
    \varepsilon = \frac {N \cdot 2r\cdot v} {A_{total}}
\end{equation}
\citep{Gurarie2012}

We then use this as the rate parameter $lambda$ to repeatedly draw the time of the next discovery event for this individual from the exponential distribution, we do this for each male from when they mature up to when the females matrure.

When an event occurs we select a nest randomly and then the male will either occupy it or contest another male for it.

%I think I might use a priority queue for this...

@@

\subsection{Occupying males}
The probability of a male abandoning a nest $j$, is a function of the  the reproductive value of the nest $R_j$ and the abandon trait of the individual $\psi_i$:

assumption: $1 < R_j$,  $1 < \psi_i$

\begin{equation} \label{eq:abandon}
    Pr({\rm abandon} | R_j ) = \frac{1}{1 + \psi_i \cdot R_j}
\end{equation}

@@

    \item Aggressiveness, influences escalation in contests
    Trade offs:
    \begin{itemize}
        \item more likely to engage in fights with a low probability of winning
    \end{itemize}

@@

\subsection{Algorithms}
\subsection{Event Algorithms}
So far we are considering two algorithms for stepping through events one is a stochastic simulation algorithm and a simpler (although more computationally expensive) naive stepping algorithm.
\subsection{SSA}
Simulating the population dynamics requires accounting for four types of events:

An efficient method for stepping the population is to step between successive events. So rather than stepping the system over a given time interval and considering any transitions that may occur in that time period, the system is stepped between successive events by drawing the next event and the time at which the event occurs from appropriate distributions.  We can identify time to next event using Gillespie minimal process method \citep{Gillespie-1976}, which indicates that with a given event rate, the distribution of waiting times follows an exponential distribution.

In the SSA algorithm \citep{Gillespie-1976} there are 6 different states
\begin{enumerate}
    \item $S_1 = $ immature male
    \item $S_2 = $ mature searching male
    \item $S_3 = $ unoccupied nest
    \item $S_4 = $ occupied nest
    \item $S_5 = $ recently defeated male
    \item $S_6 = $ recently changed nest
\end{enumerate}

these states have $R_\mu$ different interactions
\begin{table}[h!]
    \caption{state changes}
    \centering
    \begin{tabular}{c | c | c | c | l }
        \hline
            Event Number & reactants  & & products & Description\\
        \hline
        \hline
            $R_1$       &   $S_1$       & $--c_1-->$      &   $ S_2$      & maturation\\
            $R_2$       &   $S_2 + S_3$ & $--c_2-->$      &   $S_4 $      & occupation\\
            $R_3$       &   $S_2 + S_4$ & $--c_3-->$      &   $S_4 + S_5$ & contest\\
            $R_4$       &   $S_5$       & $--c_4-->$      &   $S_2$       & recovery\\
            $R_5$       &   $S_2 + S_4$ & $--c_5-->$      &   $S_4$       & contest with death\\
            $R_6$       &   $S_4$       & $--c_6-->$      &   $S_2 + S_3$ & abandon event\\
            $R_7$       &   $S_1$       & $--c_7-->$      &   $0$         & metabolic / predation event\\
            $R_8$       &   $S_2$       & $--c_8-->$      &   $0$         & predation event\\
            $R_9$       &   $S_4$       & $--c_9-->$      &   $S_3$       & metabolic / predation event\\
            $R_{10}$    &   $S_6$       & $--c_{10}-->$   &   $S_4$       & nest handover\\
        \hline
    \end{tabular}

\end{table}

$c_n$ is called the stochastic reaction constant\\
$c_n \cdot dt = $ the average probability that a particular combination of $R_n$, reactant molecules will react accordingly in the next infinitesimal time interval $dt$\\
s
\begin{verbatim}
I have the probability of encountering a   
\end{verbatim}

\citep{Gillespie-1976}

\clearpage
\subsection{Naive Event Algorithm}
we have 3 lists
\begin{verbatim}
    immature males
    searching males
    nests
\end{verbatim}

\begin{verbatim}
time = 0
while time < time when the females mature:
    check which males have matured
        add them to the list of searching males
    for each occupied nest:
        deduct metabolic costs from the occupier
        remove dead males

    for each searching male m:
        deduct metabolic costs from m
        remove dead males
        reduce time until next event for m
        if time until next event <= 0:
            reset the time until next event
            if random > exploration_prob:
                choose a nest at random
                if it is occupied:
                    contest the occupying male for control of the nest
                    if m has won the contest:
                        m occupy's n
                        the loser is added to the list of searching males
                else:
                    occupy the nest
    t += delta_time
\end{verbatim}

\clearpage
\subsection{Contest Algorithm}
See Fig. \ref{fig:contests}.
For a defender $x_{def}$ vs an attacker $x_{atk}$, there are 3 (currently 2) stages:
each individual $i$ has an aggression trait $a$ and a mass $m$. At any point in time it also has an amount of energy $E_{current}$.  
The aggression of the individual is a function of $E_{current}$:

%not sure what's going on with this one
%\begin{equation}\label{eq:logit}
 %   logit^{-1}(\alpha) = \frac{exp(\alpha)}{1 + exp(\alpha)}
%\end{equation}

\begin{equation}\label{eq:aggresstion}
f(E_{current})=\frac{L}{1+e^{(-K(E_{current}-E_{o})}}
\end{equation}

The contests may last are escalated to different degrees with increasing energy costs for both participants.  Either participant may abandon the contest at any moment.  We assume that there are energetic costs, but that there is no time cost to contests.  Traits $K$ and $E_{o}$ are heritable and potentially under selection.

The probability of the attacker escalating is:

The first stage is the display phase.  The second phase is the fight phase.  The cost of escalating for both is:

\begin{equation}\label{eq:defender_escalating}
E_{spent}=equation you guys figured out with all the \lambda 's
\end{equation}

>>>>

We consider a population of males competing for mates. The population is one with non-overlapping generations, and within each generation there is a predefined annual cycle. In spring, up to $K$ individuals hatch from eggs, grow throughout spring and summer to increase size, and then within a short period, the entire population mates, females lay eggs and everyone dies. The short duration of mating period is such that each male and female only mates once. The population then re-establishes from eggs the following spring.

@@

\subsection{Immature phase}
% TODO update this to reflect what happens in the model
% ie pullign the maturation times from the logistic cdf with location and parameters given

Within the immature phase, we assume males grow with rate
\begin{equation} \label{eq:growth}
\frac{{\rm d} m}{{\rm d} t} = a \, m^ b
\end{equation}
where $a$ is mass-based rate of growth. Integrating eq. \ref{eq:growth}, the size of individuals at time $t$, having started growing at time $0$ is given by
\begin{equation} \label{eq:growth}
m(t) = \left(m(0)^{1-b} + t a(1-b)\right)^{\frac1{1-b}},
\end{equation}

while the time $\tau$ taken for a male to grow from $m(0) \rightarrow m$ is then
\begin{equation} \label{eq:tau}
\tau(m_0, m) = \frac1{a(1-b)}\left(m ^{1-b} - m_0 ^{1-b}\right).
\end{equation}

@@

\subsection{Searching Males}
When males mature they begin searching for a females nest.

The area that a male $m$ with speed trait $v$ and radius trait $r$ can search in one time step is
\begin{equation}
    2 \cdot r \cdot v
\end{equation}
See Fig. \ref{fig:area searched}.

@@

The encounter rate $\varepsilon$ for an individual is:
\begin{equation} \label{eq:encounter rate}
    \varepsilon = \frac {N \cdot 2 \cdot r \cdot v} {A_{total}}
\end{equation}
\citep{Gurarie2012}

We then use this as the rate parameter $lambda$ to repeatedly draw the time of the next discovery event for this individual from the exponential distribution, we do this for each male from when they mature up to when the females matrure.
When an event occurs we select a nest randomly and then the male will either occupy it or contest another male for it.

When a male is kicked out of a nest we discard the old events and generate new ones.

@@

\clearpage
\subsection{Contest Algorithm}
See Fig. \ref{fig:contests}.
For a defender $x_{def}$ vs an attacker $x_{atk}$, there are 3 (currently 2) stages:
The first stage is the display stage, the second phase is the fight stage.
each individual $i$ has mass $m$. At the time of the contest it also has an amount of energy $E_{current}$.  
The aggression of the individual is a function of $E_{current}$:

\begin{equation}\label{eq:aggression}
a_i = f(E_{current})=\frac{L}{1+e^{(-K(E_{current}-E_{o})}}
\end{equation}

$L$ is a scaling parameter.

Contests can be escalated to different degrees with increasing energy costs for both participants. 
Either participant may abandon the contest at any moment, they do this by choosing not to escalate.
We assume that there are energetic costs, but that there is no time cost to contests ie they occur instantanously.

Traits $K$ and $E_{o}$ are heritable and potentially under selection.

The probability of the attacker escalating to  is:

The energetic cost of escalating is:
\begin{equation}\label{eq:defender_escalating}
E_{spent}=equation you guys figured out with all the \lambda 's
\end{equation}

%\begin{figure}[h!]
%\centering
%\includegraphics[width=10cm,height=10cm,keepaspectratio]{figures/example_costs}
%\caption{Example of escalating energy costs,}
%\label{fig:contests costs}
%\end{figure}

@@

    \item Aggressiveness, determined by two inheritable traits $k$, $E_0$
    Trade offs:
    \begin{itemize}
        \item more likely to engage in fights with a low probability of winning
    \end{itemize}

@@
